An introduction to Docker for reproducible research and collaboration

### 2014-09-12 manuscript.Rmd
writing away

Reproducible research, and in particular, computational reproducibility,
has recieved an increasing level of attention throughout the scientific
community as all steps of the scientific process, from data collection
and processing, to analyses, visualizations and conclusions depend ever
more on computation and algorithms.  Though in principle this algorithmic
dependence should make such research easier to reproduce -- computer codes
being both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

At the heart of the computational reproducibility challenge is the emmense, 
complex, diverse, and rapidly changing nature of the computer software

It is possible, and indeed often desirable, to tackle this problem
at a much higher level than that of the operating system itself.
For instance, it is often sufficient to address these concerns at the
level of a particular scripting language rather than the level of the
operating system. Most major scripting languages have dedicated tools for
handling dependencies and versioning (e.g. Ruby gems, R packages, Python
eggs, perl modules, javascript node package manager); with some languages
having the fortune or misfortune to have multiple systems. Though systems
differ somewhat in their strategies and their effectiveness, all are
generally vulnerable to changes in their lower-level dependencies (such as
compilers, C libraries, and so forth) that may differ between platforms.
More pointedly, researchers whose work is not easily isolated to a single
such environment but combines tools from different systems (including
combining those tools used for formatting & visual display, such as
LaTeX or pandoc, with scripting or analysis tools such as R). Moreover,
addressing reproducibility at the level of the operating system avoids
constructing approaches unique to a specific language or tool.

Two dominant paradigms have emerged to address these issues so far:
workflow software and cloud computing on virtual machines.  Here I
consider how a popular emerging technology, Docker, based on Linux
containers, offers a third route that addresses pratical and theoretical
concerns for reproducible research not addressed by either of these
paradigms.

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technology discussed here, but stems rather from a
reluctance to publish the code used to generating the results in the
first place [@Barnes2010]. Despite extensive evidence to the contrary
[e.g. @Ince2012], many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, we cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

Yet this coin has two sides: By lowering the barriers
these barrier to re-use through the appropriate infrastructure first, the research
community as a whole may be better positioned to take advantage of code 
sharing practices when they are adopted.  This has cultural implications as well:
many researchers may be reluctant to publish code today because they fear
a primarily one-way street: others will have the technical savvy to benefit
from their hard work, while they may not benefit from the work produced by others.
Lowering the technical barriers to reuse provides immediate practical benefits
that make this exchanged into a more balanced, two-way street.

The issue most domain researchers do not recognize as an issue.  If computational reprod

Before we begin, it is important to distinguish between the nature of the problem we seek to address here and other elements of computationally reproducible research.  While this approach may address some of the same challenges as workflow software, dynamic documents / executable papers and ad hoc packaging systems, it can always been seen to supplement rather than replace such solutions.  A good example comes from dynamic documents, which seek to embed the code required to re-generate the results within the manuscript itself.

http://doi.ieeecomputersociety.org/10.1109/MCSE.2012.62

@@

# Typical solution approaches

- Documentation
- Standard environments
- Virtual machines & Cloud computing

# Challenges with common approaches

# Emerging approaches: DevOpts

- DevOpts: Scripting your environment
- Vagrant

@@

My original motivation was more on the research side, primarily from NESCent's reproducibility study and Rich's reproducibility ropensci blog post, where you might recall Docker being discussed in the comments.

Hilmar highlighted 3 reasons why reproducibility efforts fail even for papers that seem to do the right things: 1) dependency hell frustrating installation, 2) small holes in documentation can be major barriers for "novices" (even if novices are experts in nearby fields and nearby languages), 3) software dependency evolution breaks existing code (code rot).

Rich highlighted about 4 reasons: 1) dependency hell / incomplete documentation on versions, 2) all the scripts & software for installing managing the 'reproducibility' part of their analysis (travis, etc) were as complex as the analysis itself. (more than half the software dependencies were things like RCurl,  knitr & packrat etc).  3) Running the analysis on Travis CI means it has to finish in 50 minutes, 4) the learning curve for Github + Travis + etc is prohibitive.

Docker provides a "DevOpts" approach, replacing user documentation on how to install with scripts that handle provisioning.  This pretty much solves Hilmar's #1 & #2.  You can maintain a timestamped binary of the Docker image to address Hilmar's #3.

Rich's #2 is largely handled by providing an image where this stuff is all set up for you, so it building on the ropensci docker image, you need to add only what is necessary for your actual analysis.  His #3 is addressed bc instead of running on Travis, you can run this locally (on Windows/Mac/Linux) or on a cloud of your choice with as much horsepower as you need.  I think this lowers the learning curve (#4) some, (for instance, you can use git in the usual way but you can also fully interact without knowing git).

My hope is, that instead of the rather giant set-up Rich et al needed, a user could just have their paper as say, an R script or rmarkdown file, develop & run it locally in the docker-ropensci environment, and then anyone having trouble getting it to run in their native R should at least be able to run it in the docker-ropensci environment.  Docker is super-easy to extend: they could make their own minimal Dockerfile like this: https://github.com/ropensci/RNeXML/blob/devel/manuscripts/Dockerfile There's a few other cool things about Docker (The git-like way it handles images etc) too.

>>>>

Reproducible research has recieved an increasing level of attention
throughout the scientific community [The Economist 2014]. A all steps
of the scientific process, from data collection and processing, to
analyses, visualizations and conclusions depend ever more on computation
and algorithms, _computational reproducibility_ has recieved particular
attention [Merali 2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

## A technical and cultural problem

It is worth observing from the outset that the primary barrier to
computational reproducibility in many domain sciences has nothing
to do with the technological approaches discussed here, but stems rather from a
reluctance to publish the code used to generating the results in the
first place ([Barnes 2010]). Despite extensive evidence to the contrary
(e.g. [Ince 2012]), many researchers and journals continue to assume that
summary descriptions or pseudo-code provide a sufficient description of
computational methods used in data gathering, processing, simulation,
visualization, or analysis.  Until such code is available in the first
place, we cannot even begin to encounter the problems that the approaches
discussed here set out to solve. As a result, few domain researchers
may be fully aware of the challenges involved in effectively re-using
published code.

A lack of requirements or incentives no doubt plays a crucial role
in discouraging sharing ([Barnes 2010]). Nevertheless, it is easy to
underestimate the siginficant barriers raised by a lack of familiar,
intuitive, and widely adopted tools for addressing the challenges of
computational reproducibility. Surveys and case studies find that a lack
of time, more than innate opposition to sharing, discourages researchers
from providing code ([Stodden 2010], [FitzJohn 2014]).

<!-- Time savings. Personal reproducibility. Collaboration. Teaching. -->

<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents

<!-- Operating system - level approach -->
At the heart of the computational reproducibility challenge is the 
complexity, diversity, and rapidly changing nature of the computer software
that underpin such work.

<!-- Current approaches at the Operating Systems level -->

Two dominant paradigms have emerged to address these issues so far:
workflow software (e.g. [Atintas 2004], [Hull 2006])  and
cloud computing on virtual machines (e.g. [Dudley 2013], [Howe 2012]). 
Workflow software provides very elegant technical solutions to the
challenges of communication between diverse software tools, capturing
provenance in graphically driven interfaces, and handling issues from of
versioning dependencies to data access.  Workflow solutions are often 
built by well-funded collaborations between domain scientists and computer
scientists, and can be very successful in the communities within which 
they recieve substatianal adoption. Nonetheless, most workflow systems 
struggle with relatively low total adoption overall ([Gin 2007, Dudley 2013]).

[Dudley 2013] gives several reasons for this:

> (i) efforts are not rewarded by the current academic research and
funding environment; (ii) commercial software vendors tend to protect
their markets through proprietary formats and interfaces; (iii)
investigators naturally tend to want to ‘own’ and control their
research tools; (iv) even the most generalized software will not be
able to meet the needs of every researcher in a field; and finally (v)
the need to derive and publish results as quickly as possible precludes
the often slower standards-based development path.

In short, workflow software expects a new approach to computational research

In contrast, virtual machines (VMs) offer a far more brute-force approach.  If the goal
is to reproduce everything as it ran on the researcher's own computer, well 
then we can just take a snapshot of the entire computer: data, software, 
operating system and all.

To make this practical, [Dudley 2013] and [Howe 2012]
propose using virtual machine images that will run on the cloud, such as Amazon's
EC2 system, which is already based upon this kind of virtualization.

Critics of the use of VMs to support reproducibility highlight that the 
approach is too much of a black box and ill suited for

Here I consider how a popular emerging technology, Docker, based on
Linux containers, offers a third route that addresses pratical and
theoretical concerns for reproducible research not addressed by either
of these paradigms.

@@

## Alternatives and other considerations

<!-- Package managers -->

It is possible, and indeed often desirable, to tackle this problem
at a much higher level than that of the operating system itself (e.g. [Ooms 2014]).
For instance, it is often sufficient to address these concerns at the
level of a particular scripting language rather than the level of the
operating system. Most major scripting languages have dedicated tools for
handling dependencies and versioning (e.g. Ruby gems, R packages, Python
eggs, perl modules, javascript node package manager); with some languages
having the fortune or misfortune to have multiple systems. Though systems
differ somewhat in their strategies and their effectiveness, all are
generally vulnerable to changes in their lower-level dependencies (such as
compilers, C libraries, and so forth) that may differ between platforms.
More pointedly, researchers whose work is not easily isolated to a single
such environment but combines tools from different systems (including
combining those tools used for formatting & visual display, such as
LaTeX or pandoc, with scripting or analysis tools such as R). Moreover,
addressing reproducibility at the level of the operating system avoids
constructing approaches unique to a specific language or tool.

<!-- Synergistic with other reproducible-research technology -->

The approaches discussed here should be seen as largely synergistic
with, rather than an alternative to, other topics in computationally
reproducible research. For example, consider the dynamic documents
approach ([Leisch 2002], [Peng 2011], [Xie 2013]), which seek to embed
the code required to re-generate the results within the manuscript
itself. This process serves as a way to distribute computational
code with publications, to ensure that methodological discriptions,
parameter values, figures, and so forth remain synchronized with the code,
and to provide a way for other researchers to both tweak the code and
'execute' the paper to reproduce the results.  While these features
are very desirable from the perspective of reproducible research,
they inevitably increase the number of software dependencies involved,
including those tools that support combining code with text (e.g. [Xie
2013]) as well as tools merely for the handling the formating and layout
(e.g. LaTeX, pandoc).  Consequently, such a paper can fail to execute
successfully for reasons having nothing to do with the computations
involved in the research, but rather due to software responsible for
implementing these dynamic features.  This problem is largely resolved by
providing a consistent development environmment, making the approaches
discussed here a natural complement to that of dynamic documents.
A similar argument can be made regarding the use of workflow systems,
which can introduce additional softare dependencies even as they seek
to make it easier to work across them.

<!-- Cultural sharing practices -->
We noted at the outset that cultural expecations responsible for a
lack of code sharing practices in many fields are a far more extensive
primary barrier to reproducibility than the technical barriers we have
discussed here. Nevertheless, it may be worth considering how solutions
to these technical barriers can influence the cultural landscape as well.
Many researchers may be reluctant to publish code today because they
fear a it will be primarily a one-way street: more technical savvy
researchers then themselves can benefit from their hard work, while
they may not benefit from the work produced by others.  Lowering the
technical barriers to reuse provides immediate practical benefits
that make this exchanged into a more balanced, two-way street. Another
concern is that the difficulty imposed in preparing code to be shared,
such as providing even semi-adequate documentation or support for other
users to be able to install and run it in the first place is too high
[Stodden 2010]. By lowering the barriers these barrier to re-use through
the appropriate infrastructure first, we may also reduce certain cultural
barriers to sharing.

[Vandewalle 2009]: http://doi.org/10.1109/MSP.2009.932122
[Merali 2010]: http://doi.org/10.1038/467775a "Computational Science ... error: why scientific programming does not compute"
[Gil 2007]: http://doi.org/10.1109/MC.2007.421 "Examining the Challenges of Scientific Workflows"
[Hull 2006]: http://doi.org/10.1093/nar/gkl320 "Taverna: a tool for building and running workflows of services"
[Altintas 2004]: http://doi.org/10.1109/SSDM.2004.1311241 "Kepler: an extensible system for design and execution of scientific workflows"
[Howe 2012]: http://doi.org/10.1109/MCSE.2012.62
[Dudley 2013]: http://doi.org/10.1038/nbt1110-1181 
[Clark 2014]: https://berkeley.app.box.com/s/w424gdjot3tgksidyyfl
[Ooms 2013]: http://arxiv.org/abs/1303.2140v2
[Ooms 2014]: http://arxiv.org/abs/1406.4806
[Peng 2011]: http://doi.org/10.1126/science.1213847
[Barnes 2010]: http://doi.org/10.1038/467753a
[Ince 2012]: http://doi.org/10.1038/nature10836
[Stodden 2014]: http://doi.org/10.3233/SJI-140818
[Stodden 2010]: http://doi.org/10.2139/ssrn.1550193

@@

Why computational reproducibility is difficult

Recent studies have highlighted 5 reasons

1) "Dependency Hell"
  a) Imprecise documentation. Small holes in the documentation can be major barriers for "novices" (Sensu ..., where novices may still be experts in nearby langauages). 
  b) Code rot. Dependency evolution breaks existing code

Solution to this is scripted installations: DevOpts. At the Operating System level!

2) Complexity of existing reproducibility options 
  a) Virtual Machines
  b) Workflow software
  c) Continuous Integration / software development practices
Barriers to entry / learning
Barriers to reuse / remix

Docker does 4 new things:

1) Remix.  Titus has an excellent post, "Virtual Machines Considered
Harmful for reproducibility" [1] , essentially pointing out that "you
can't install an image for every pipeline you want...".  In contrast,
Docker containers are designed to work exactly like that -- reusable
building blocks you can link together with very little overhead in
disk space or computation.  This more than anything else sets Docker
apart from the standard VM approach.

[Clark 2014]

> In scientific computing the environment was commonly managed via Makfiles & Unix-y hacks, or alternatively with monolithic software like Matlab. More recently, centralized package management has provided curated tools that work well together. But as more and more essential functionality is built out across a variety of systems and languages, the value -- and also the difﬁculty -- of coordinating multiple tools continues to increase. Whether we are producing research results or web services, it is becoming increasingly essential to set up new languages, libraries, databases, and more.

> Documentation for complex software environments is stuck between two opposing demands. To make things easier on novice users, documentation must explain details relevant to factors like different operating systems. Alternatively, to save time writing and updating documentation, developers like to abstract over such details.

## A "DevOpts" approach

[Clark 2014]

 Provisioning scripts. Docker images are not 'black boxes'. A
"Dockerfile" is a simple make-like script which installs all the
software necessary to re-create ("provision") the image.  This has
many advantages: (a) The script is much smaller and more portable than
the image. (b) the script can be version managed (c) the script gives
a human readable (instead of binary) description of what software is
installed and how. This also avoids pitfalls of traditional
documentation of dependencies that may be too vague or out-of-sync.
(d) Other users can build on, modify, or extend the script for their
own needs.  All of this is what we call the he "DevOpts" approach to
provisioning, and can be done with AMIs or other virtual machines
using tools like Ansible, Chef, or Puppet coupled with things like
Packer or Vagrant (or clever use of make and shell scripts).

For a much better overview of this "DevOpts" approach in the
reproducible research context and a gentle introduction to these
tools, I highly recommend taking a look at Clark et al [2].

3) You can run the docker container *locally*.  I think this is huge.
In my experience, most researchers do their primary development
locally.  By running RStudio-server on your laptop, it isn't necessary
for me to spin up an EC2 instance (with all the knowledge & potential
cost that requires).  By sharing directories between Docker and the
host OS, a user can still use everything they already know -- their
favorite editor, moving files around with the native OS
finder/browser, using all local configurations, etc, while still
having the code execution occur in the container where the software is
precisely specified and portable.  Whenever you need more power, you
can then deploy the image on Amazon, DigitalOcean, a bigger desktop,
your university HPC cluster, your favorite CI platform, or wherever
else you want your code to run.  (On Mac & Windows, this uses
something called boot2docker, and was not very seamless early on.  It
has gotten much better and continues to improve.)

Related challenges:  RStudio and a consistent environment. 
- Linked volumes 
- linked containers 
- signed containers?

Hub / image delivery

4) Versioned images.  In addition to version managing the Dockerfile,
the images themselves are versioned using a git-like hash system
(check out: docker commit, docker push/pull, docker history, docker
diff, etc).  They have metadata specifying the date, author, parent
image, etc.  We can roll back an image through the layers of history
of its construction, then build off an earlier layer.  This also
allows docker to do all sorts of clever things, like avoiding
downloading redundant software layers from the docker hub.  (If you
pull a bunch of images that all build on ubuntu, you don't get n
copies of ubuntu you have to download and store).  Oh yeah, and
hosting your images on Docker hub is free (no need to pay for an S3
bucket... for now?) and supports automated builds based on your
dockerfiles, which acts as a kind of CI for your environment.
Versioning and diff'ing images is a rather nice reproducibility
feature.

5) Linking images 
Docker is not a virtual machine; containers are designed expressly to be remixable blocks.  You can put an R engine on one container and a mysql database on another and connect them. Docker philosophy aims at one function per container to maximize this reuse.  of course it's up to you to build this way rather than a single monolithic dockerfile, but the idea of linking containers is a technical concept at the heart of docker that offers a second and very different way to address the 'remix' problem of VMs.

Cultural challenges
